---
title: "Dataset and AMR phenotype characterization"
output:
  md_document: default
---

### Objective 

This R notebook contains the code used to analyze:
  - diverse meta information associated with the different strains used in this study (all strains but the ECOR72), these metadata include data associated with sample information (isolation country, year, host...), sequencing information (NGS technology, accession numbers, quality, etc ...)
    along with genetic information such as number of bp, number of coding sequences, etc...
  - the antimicrobial resistance (AMR) phenotypes in the dataset

---

```{r}
library(tidyverse)
library(ggwordcloud)
library(arcadiathemeR)
library(maps)

```

--- 

# Meta-information analysis of the cohort

## Data import

```{r}
 ## Samples list for which we obtained genome sequencing files from SRA accession number
sample_list=read.csv('dataset_generation/data/sample_list_SRA.csv')

 ## Genome information from BVBRC
genome_data=read.csv('dataset_analysis/data/dataset_analysis/BVBRC_genome_May31.csv')

```

## Selection of the strains used in this study

```{r}

# 1- Extracting the Genome Names in the BV-BRC dataset associated with the SRA accession number we used to obtain the sequencing files

samples=c(unique(sample_list$samples))
genome_samples=subset(genome_data, genome_data$SRA.Accession%in%samples)

   ## Some Genome Names are associated with multiple SRA accession number and we need to treat them separately 

out=setdiff(samples, genome_samples$SRA.Accession)
pattern <- paste(out, collapse="|")
matching_rows <- grep(pattern, genome_data$SRA.Accession)
subset_df <- genome_data[matching_rows, ]

   ## And combine them 

genome_samples=rbind(genome_samples,subset_df) # dataframe with all meta-information of the strains used in this study (excluding the ECOR72 strains)

```

## Selection of metadata of interest

The metadata we want to investigate are:
  - genome size (number of bp)
  - number of CDS
  - collection year
  - isolation country
  - host
  
```{r}

metadata=genome_samples %>%
  select(Genome.Name, Size, CDS, Collection.Year, Isolation.Country, Host.Name)

empty_counts <- colSums(metadata == "")

empty_counts # indicates for each metadata type the number of strains for which the information is missing

```
### Genome size

```{r}
min_size=min(metadata$Size)
max_size=max(metadata$Size)
mean_size=mean(metadata$Size)
median_size=median(metadata$Size)

size_plot <- ggplot(metadata, aes(x = Size)) +
  geom_histogram(binwidth = 50000, fill='#7A77AB',col='#484B50')+
  scale_x_continuous(limits = c(4000000, 6000000),
                     breaks = seq(4000000, 6000000, by = 500000),
                     labels = function(x) round(x / 1000000, 1)) +
  theme_arcadia(x_axis_type = "numerical") +
  xlab('Genome size (millions base pair)')+
  ylab('Number of strains')

min_size
max_size
mean_size
median_size

size_plot

```
### CDS number
```{r}

min_cds=min(metadata$CDS)
max_cds=max(metadata$CDS)
mean_cds=mean(metadata$CDS)

min_cds
max_cds
mean_cds

```

### Collection year
```{r}
year_counts <- table(metadata$Collection.Year)
year_counts

```

### Host
```{r}
hosts_counts <- as.data.frame(table(metadata$Host.Name))
hosts_counts

```

### Country
```{r}
country_counts <- as.data.frame(table(metadata$Isolation.Country))
colnames(country_counts)=c('country','count')

#Reformatting the data to use them with the map package and represent the data on a world map. This includes removing the strains without country information and consolidating strains from the United Kingdom and England into a single count.

country_for_map=country_counts[-1,]
country_for_map <- data.frame(
  country = c("United Kingdom", setdiff(country_for_map$country, c("England", "United Kingdom"))),
  count = c(sum(country_for_map$count[country_for_map$country %in% c("England", "United Kingdom")]), 
            country_for_map$count[!country_for_map$country %in% c("England", "United Kingdom")])
)

# Changing United Kingdom to UK to match the country name in the map package
country_for_map$country <- gsub("United Kingdom", "UK", country_for_map$country)

# Mapping the data onto a world map
## Generating the data
world_map <- map_data("world")
merged_data <- merge(world_map, country_for_map, by.x = "region", by.y = "country", all.x = TRUE)

# Adding a column that categorizes the countries based on how many strains have been sampled there

merged_data <- merged_data %>%
  mutate(cat = case_when(
    count > 500 ~ "More than 500 strains",
    count < 500 ~ "Less than 500 strains",
    is.na(count) ~ NA_character_
  ))


# Plot the world map with ggplot2
ggplot() +
  geom_map(data = merged_data, map = world_map,
           aes(x = long, y = lat, map_id = region, fill = cat),
           color = "white", size = 0.1) +
  scale_fill_manual(values = c("> 500 strains" = "#F28360", "< 500 strains" = "#5088C5", "NA" = "#EBEDE8"),
                    na.value = "grey90", name = "Number of strains sampled from a country") +
  theme_arcadia() +
  theme(legend.position = "bottom") 



```

---

# AMR phenotype analysis

In this part we investigate the AMR information available for the strains studied in the Pub.

## Data import

```{r}
amr_data=read.csv('dataset_analysis/data/dataset_analysis/BVBRC_genome_amr_May31.csv')

# Extracting antibiotic info for the strains using their genomes names
amr_data_sub=subset(amr_data, amr_data$Genome.Name%in%genome_samples$Genome.Name)

#Correspondence antibiotic - antibiotic class
antibiotic_class=read.csv('dataset_analysis/data/dataset_analysis/Antibiotic_class.csv')

```

## Reformatting the data, removing inconsistent information
For some strains, multiple AMR phenotypes were reported for the same antibiotic, when these AMR phenotypes are inconsistent we remove them.

```{r}
     
# Keeping consistent repetitive AMR phenotype but keeping only a single row entry

amr_data_sub_short=amr_data_sub%>% select(Genome.Name, Antibiotic, Resistant.Phenotype)%>%
  filter(Resistant.Phenotype != "")  %>% distinct()

# Removing any inconsistent AMR phenotype for the same strain and antibiotic

amr_data_filtered <- amr_data_sub_short %>%
  group_by(Genome.Name, Antibiotic) %>%
  filter(n_distinct(Resistant.Phenotype) == 1) %>%
  ungroup()

```

## Generating the phenotype matrix
The phenotype matrix provides AMR information for all strains (rows) and antibiotics (columns)
NAs indicate that no AMR information available

```{r}
  # Preparing phenotype matrix

amr_data_matrix=amr_data_filtered%>%
  spread(Antibiotic, Resistant.Phenotype)

#write.csv(amr_data_matrix,"dataset_analysis/results/phenotype_matrix_08302024.csv")

```

## Data exploration
In this part, we explore and visualize different information related to AMR phenotypes

### Number of Antibiotics or Combinations of antibiotics
```{r}
antibiotics=c(unique(amr_data_filtered$Antibiotic)) # 50 antibiotics with at least 1 known phenotype
length(antibiotics)
```

## Number of known AMR phenotypes per strain

```{r}
# For each strains counts how many AMR phenotypes are available
count_sample=amr_data_filtered%>%group_by(Genome.Name)%>%
  summarize(count=n())

head(count_sample,5)

# Visualization of the distribution of known AMR phenotype per strain
amrphen_per_strain=ggplot(count_sample, aes(x=count))+
  geom_histogram(bins=max(count_sample$count), fill='#97CD78', col='#484B50') + 
  xlab('Number of known AMR phenotypes in a strain') +
  ylab ('Number of strains')+
  theme_arcadia(x_axis_type = "numerical")

amrphen_per_strain

```

## Number of strains with known AMR phenotype per antibiotic

```{r}
# For each antibiotic we look at how many strains have available AMR phenotype information
count_antibio=amr_data_filtered%>%group_by(Antibiotic)%>%
  summarize(count=n())

head(count_antibio,10)

amrphen_per_antibiotic=ggplot(count_antibio, aes(x=count))+
  geom_histogram(binwidth =100, fill='#C85152', col='#484B50') + 
  xlab('# of knwown AMR phenotypes') +
  ylab ('# of antibiotics')+
  theme_arcadia(x_axis_type = "numerical")

amrphen_per_antibiotic


```

## Antibiotics with 500 or more AMR phenotypes available: analysis of AMR distribution
```{r}
# Selecting antibiotics with more than 500 AMR phenotypes 
list_antibio_high <- count_antibio %>%
  filter(count >= 500) %>%
  pull(Antibiotic) %>%
  unique()

df_antibio_high=subset(amr_data_filtered, amr_data_filtered$Antibiotic%in%list_antibio_high)

# For each antibiotic counts how many 'Resistant, 'Susceptible' and 'Intermediary' phenotype
phenotypes_counts <- df_antibio_high %>%
  group_by(Antibiotic, Resistant.Phenotype) %>%
  summarise(Count = n(), .groups = 'drop')

phenotypes_counts_plot=phenotypes_counts%>%
  dplyr::select(Antibiotic, Resistant.Phenotype, Count, Class)%>%
  mutate(Antibiotic = factor(Antibiotic, levels = unique(Antibiotic[order(Class)])))

phenotypes_counts_plot$Resistant.Phenotype <- factor(phenotypes_counts_plot$Resistant.Phenotype, levels = c("Susceptible", "Intermediate","Resistant" ))


phen_distribution=ggplot(phenotypes_counts_plot, aes(x=Antibiotic, y=Count, fill=Resistant.Phenotype))+
  geom_bar(stat='identity') + 
  xlab('Antibiotic') + ylab('number of strains') +
  theme_arcadia(x_axis_type = 'categorical')+
  scale_fill_arcadia(palette_name = "secondary", reverse = TRUE, name='AMR phenotype')+
  theme(axis.text.x = element_text(angle=90,hjust=1))+
 theme(
    legend.position = "bottom",
    legend.box = "vertical" ,
    legend.margin = margin(t = 25))+
  guides(fill = guide_legend(order = 1, title.position = "top",ncol = 1))

phen_distribution
```

## Resistant strains

```{r}

Resistant=subset(amr_data_filtered, amr_data_filtered$Resistant.Phenotype=='Resistant')

list_antibio_resist=c(unique(Resistant$Antibiotic))

#Number of resistant AMR per strain

res_count=Resistant%>%group_by(Genome.Name)%>%
  summarize(count=n())

head(res_count)

#Resistance per antibiotic class

res_class=Resistant %>% left_join(antibiotic_class, by="Antibiotic")%>%
  group_by(Class)%>%
  summarize(count=n())

res_per_class=ggplot(res_class, aes(label=Class,size=count, color=Class))+
  geom_text_wordcloud() +
  scale_size_area(max_size = 10)+
  theme_arcadia() +
  scale_color_arcadia(palette_name = "primary")

res_per_class

```

## Multi-drug resistant strains
Here we define Multi-drug resistant bacteria if they are resistant to 3 or more different class of antibiotic

```{r}

MDR_candidate=res_count %>%
  filter(count >= 3) %>%
  pull(Genome.Name) %>%
  unique()

MDR_candidate_DF=subset(Resistant, Resistant$Genome.Name%in%MDR_candidate)
MDR_candidate_DF=left_join(MDR_candidate_DF, antibiotic_class, by="Antibiotic")

strains=c(unique(MDR_candidate_DF$Genome.Name))

length(strains)

```


```{r}
sessionInfo()
```